---
title: "Reproducible Research"
author: "Sylvia Wotus"
output: html_document
---

#### Loading and preprocessing the data

```{r,echo=TRUE}
data <- read.csv("activity.csv")
head(data)
summary(data)
str(data)
```

#### The mean total number of steps taken per day

```{r,echo=TRUE}
library(ggplot2)
dev.off() #in case some previous plot commands are still active
stepsDay <- tapply(data$steps, data$date, FUN=sum, na.rm=TRUE)

```

```{r,echo=TRUE}
qplot(stepsDay, binwidth=1000, xlab="Total number of steps taken per day") +
        geom_vline(xintercept = c(mean(stepsDay, na.rm=TRUE)), colour = "red", size=1) + 
        geom_vline(xintercept = c(median(stepsDay, na.rm=TRUE)), colour = "blue", size=1)
```
On the graph, the median is represented by the red vertical line and the mean by the blue one

```{r,echo=TRUE}
mean(stepsDay, na.rm=TRUE)
median(stepsDay, na.rm=TRUE)

cat("The median is", median(stepsDay, na.rm=TRUE), "while the mean is", mean(stepsDay, na.rm=TRUE), ".\n" )
```

#### The average daily activity pattern

```{r,echo=TRUE}
library(ggplot2)
# aggredate function: splits the data into subsets, computes summary statistics for each, and returns the result in a convenient form.

averageStep<- aggregate(x=list(steps=data$steps),
                        by=list(interval=data$interval),FUN=mean, na.rm=TRUE)

averageStep[which.max(averageStep$steps),]

ggplot(data=averageStep, aes(x=interval, y=steps)) + geom_line() +
xlab("5-minute interval") + ylab("Average number of steps taken") + 
geom_vline(xintercept = c(835), colour = "red", size=1)

```

On the graph, the maximum of steps is represented by the red vertical line, on the interval 835.

#### Imputing missing values

```{r,echo=TRUE}
missing <-is.na(data$steps)
sum(missing)

cat("The total missing values for the Step variable are", sum(missing), "which represent", round(((sum(missing)/nrow(data))*100), 2), "% of the entire data set", ".\n")

# Function to file NA
fillNA <- function(steps, interval) 
{ filled <- NA
        if  (!is.na(steps))
        filled <- c(steps)
                else 
        filled <- (averageStep[averageStep$interval==interval, "steps"])
        return(filled)
}

updateData <- data
updateData$steps <- mapply(FUN = fillNA, updateData$steps, updateData$interval)

totalSteps <- tapply(FUN=sum, updateData$steps, updateData$date)
qplot(totalSteps, binwidth=1000, xlab="Total number of steps taken per day")

cat("The median is", median(totalSteps, na.rm=TRUE), "while the mean is also", mean(totalSteps, na.rm=TRUE), ".\n" )
```

#### Weekdays vs Weekends 

```{r,echo=TRUE}
weekdayVSweekend <- function(date) {
        day <- weekdays(date)
        if (day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
        return("weekday")
        else if (day %in% c("Saturday", "Sunday"))
        return("weekend")
        else
        stop("invalid")
}
updateData$date <- as.Date(updateData$date)

updateData$day <- sapply(updateData$date, FUN=weekdayVSweekend)
averages <- aggregate(steps ~ interval + day, data=updateData, mean)
ggplot(averages, aes(interval, steps)) + geom_line() + facet_grid(day ~ .) +
xlab("5 min interval") + ylab("Number of steps")

```

#### Difference between Weekdays and Weekends'days

```{r,echo=TRUE}
dataDifference<-data.frame(averages$interval)
dataDifference$weekday<-averages$steps[1:288]
dataDifference$weekend<-averages$steps[289:576]
dataDifference$difference<- dataDifference$weekday-dataDifference$weekend

dataDifference$threshold<-factor(ifelse(dataDifference$difference>0, 1,
                ifelse(dataDifference$difference<0,-1,0)))

ggplot(dataDifference, aes(averages.interval, difference), colour=threshold) + geom_line() + xlab("5 min interval") + ylab("Difference of steps between weekdays and weekends") + geom_hline(yintercept =0, size = 1, linetype=2)
```

